The role of cognateness in native spoken word recognition

### 2024-11-15 manuscript/manuscript.tex
Fix epreds

When some German speakers listened to the song \emph{The Power} by SNAP!
(\emph{Coyote Ugly}, 2000), many of them misheard the line ``I've got
the power'' as \emph{Agatha Bauer!} When Michael Jackson vocalised
``ma-ma-coo-sah'' in his song \emph{Wanna Be Startin' Somethin'}
(\emph{Thriller}, 1982), some Dutch speakers understood \emph{mama
appelsap} {[}mama applejuice{]}. When Spanish speakers listen to the
line ``Circumvent your thick ego'' from the song \emph{Pictures} by
System of a Down (\emph{Steal this Album!}, 2002), they tend to hear
\emph{Sácame de aquí, King-Kong} {[}take me out of here, King-Kong{]}.
When Japanese speakers listen to the line ``I want to hold your hand''
in the homonym song by The Beatles (\emph{Meet The Beatles}, 1964), they
often mishear \emph{Aho na hōnyōhan} {[}stupid urinator{]}. Outrageous
as these examples might sound to speakers of other languages, many
readers may know a few cases in their own native language. This auditory
illusion can feel quite real, and often inevitable after the first time
it happens. In Japanese, this phenomenon takes the name \emph{Soramimi}
(lit. ``empty ear''), after the \emph{Soramimi Hour}---a section of the
program \emph{Tamori Club} hosted by comedian Morita Kazuyoshi
(Tamori)---in which instances of this illusion were presented with
comedy purposes. Soramimi is a particular case of \emph{homophonic
translation}: words or phrases in one language are translated into
similar-sounding words and phrases in a different language, without
necessarily preserving the meaning
(\citeproc{ref-gasparov2006semen}{Gasparov, 2006}). This phenomenon
extends beyond song lyrics, and may occur anytime when listening to
non-native speech. For instance, when listening to ``You know me?''
(\textipa{/ju: n@U mi:}/) in English, Japanese-native listeners tend to
understand \emph{yunomi} (/\textipa{jUnomi}/), which is a type of
Japanese teacup. Overall, the occurrence of Soramimi---and more
generally, homophonic translations---suggests that speech in an
unfamiliar language has the potential to activate lexical
representations in the native language.

Although homophonic translations do not necessarily take place with the
preservation of meaning---which often results in outrageous
mistranslations like the ones aforementioned---homophonic translations
sometimes may lead to correct translations. For example, imagine a
Spanish native with no previous familiarity with any other language
listening to French for the first time. This listener may encounter the
word /\textipa{"pOKt}/ (\emph{porte}), translation of door in French.
When comparing the phonemic transcription of the French to that of its
Spanish translation /\textipa{"pweR.ta}/ (\emph{puerta}), one may be led
to infer that these two words do not have much overlap. But one thing to
note is that the voiced uvular fricative consonant /\textipa{K}/ and the
open-mid back rounded vowel /\textipa{O}/, which do not exist in Spanish
as phonemic categories, can be perceived as allophonic variations the
native (Spanish) phonemes /\textipa{R}/ and /o/, as described by
assimilation models of phoneme perception
(\citeproc{ref-best1988examination}{Best et al., 1988},
\citeproc{ref-best2001discrimination}{2001}). Although this phonetic
mismatch between the non-native word and its translation can have a
noticeable toll on comprehension Cutler et al.
(\citeproc{ref-cutler2004patterns}{2004}), the Spanish listener may
still be able to activate the phonological representation of
/\textipa{"pweR.ta}/ (\emph{puerta}) from hearing the word-form
/\textipa{"pOKt}/ (\emph{porte}), ultimately allowing them to access the
correct semantic representation. Given enough phonological similarity
between the non-native word and its translation in the native language,
a naïve listener may succeed at word comprehension when listening to an
unfamiliar language. Such phonological similarity between translation
equivalents---known as \emph{cognateness}---is common across many
languages, and is often due to typological closeness and/or
socio-historical events involving the speakers of these languages (e.g.,
migration, social contact). For example, Romance languages such as
Spanish and Catalan share many form-similar translation equivalents
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and
Catalan, respectively). Given no prior knowledge of either language, a
Spanish native speaker is likely to be much more successful at correctly
translating Catalan \emph{porta} than English \emph{door}, due to the
phonological and orthographic similarity of the former to the equivalent
in their native language.

While Soramimi (and more generally homophonic translations) are popular
and frequent phenomena, the psycholinguistic literature on the topic is
relatively scarce, and has mostly focused on the phonological variables
involved in the occurence of Soramimi. For instance, Otake
(\citeproc{ref-otake2007interlingual}{2007}) analysed 194 instances of
Soramimi, broadcasted between 1992 and 2007 by the TV show Soramimi
hour. These instances consisted of homophonic translations of English
song lyrics to Japanese words and phrases. The vast majority of the
reported instances were phrases (96\%) and a few, to single words (4\%).
The phonetic features of the presented English lyrics were preserved
with relatively high variability, with some Japanese resulting words or
phrases sharing high overlap with the original English strings of
sounds, and some sharing very little overlap. When analysing the few
instances of homophonic translations of single words, the author
identified three phonological processes that might explain how Japanese
listeners reconstructed the English input to generate the Japanese
words: insertion (e.g., \emph{cry} /\textipa{kraI}/ to \emph{kurai},
/\textipa{kuRai}/ {[}dark{]}), deletion (e.g., \emph{go},
/\textipa{goU}/ to \emph{go}, /go/ {[}go{]}, and alternation (e.g.,
\emph{low}, /\textipa{loU}/ to \emph{rou}, /\textipa{Roo}/ {[}wax{]}).
This suggests that the reported homophonic translations can be
explained, to some extent, as the result of the Japanese listeners
accommodating the strings of English sounds to the Japanese phoneme
inventory and Japanese phonotactics
(\citeproc{ref-dupoux1999epenthetic}{Dupoux et al., 1999};
\citeproc{ref-peperkamp2008perceptual}{Peperkamp et al., 2008}). Native
prosody also constrains the occurence of homophonic translations. For
instance, Kentner (\citeproc{ref-kentner2015rhythmic}{2015}) presented
in the auditory modality French-native and German-native participants
with English songs with lyrics. Homophonic translations followed
language-specific segmentation strategies. Most homophonic translations
reported by German-native participants resulted from inserting phrase
and word boundaries \emph{before} stressed syllables (following the more
frequent trochaic patter in German). On the contrary, most homophonic
translations reported by French-native participants resulted from
inserting phrase and word boundaries \emph{after} stressed syllables,
following the more frequent iambic pattern in French.

These studies have addressed the phonological aspects of homophonic
translation, but little attention has been paid to the dynamics of
lexical activation and competition that underlay this phenomenon. In the
present work, we investigated the interplay between phonological
similarity and phonological neighbourhood density in a non-native speech
perception task. In particular, we studied how the presence of
phonological neighbours in the native language lexicon affects the
extent to which listeners are able to activate the correct translation
of an auditorily presented word in thier native language. We used a
translation elicitation task in which participants listened to words
from an unfamiliar language, and then tried to guess their translation
in their native language. We will henceforth refer to the
auditory-presented words heard by participants on each trial as
\emph{presented words}, and the correct translation for the presented
words as \emph{target translations}. We explored listeners' reliance on
phonological similarity to succeed in the task by manipulating the
amount of phonological similarity between the presented words and their
target translations. Since participants were unfamiliar with the
presented language, they could only use phonological similarity between
the presented language and their native language to successfully
translate the words. We therefore predicted that participants'
performance should increase when the translation pairs are
phonologically more similar. We also predicted that there is a minimum
threshold of phonological similarity to be sufficient for translation.

We further investigated the effect of phonological competitors. Even if
the presented word and its target translation share high phonological
similarity, participants may still provide incorrect translations if the
presented word also shares high phonological similarity with other words
in the native language. Words with denser phonological neighbourhoods
are recognised more slowly and less accurately than words with sparser
phonological neighbourhoods (\citeproc{ref-dufour2003lexical}{Dufour \&
Peereman, 2003}; \citeproc{ref-goldinger1989priming}{Goldinger et al.,
1989}; \citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek,
1996}; \citeproc{ref-luce1990similarity}{Luce et al., 1990};
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). This is
especially true if such neighbours share higher phonological similarity
with the presented word, or are lexically more frequent than the target
translation (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}).
Since the interference effects of phonological neighbourhood density has
also been reported across languages
(\citeproc{ref-weber2004lexical}{Weber \& Cutler, 2004}), we
hypothesised that the size of the facilitatory role of phonological
similarity would be inversely proportional to the amount of
higher-frequency phonological neighbours of the presented word in the
target language. In the case of non-native speech recognition, the
presence of cross-linguistic pairs which are phonologically similar but
differ in meaning (e.g., false friends) may act as distractors during
lexical access, obstructing the selection of the appropriate target
translation in the listener's lexicon. For instance, Otwinowska \&
Szewczyk (\citeproc{ref-otwinowska2019more}{2019}) reported that false
friends were disadvantaged relative to non-cognates by Polish second
language learners, in contrast to cognates which were known better. It
is therefore important to investigate the joint effect of both cognates
and false friends when investigating the effect cross-linguistic
phonological similarity has on word recognition in a foreign language.
This could shed light on available strategies and challenges associated
with the early stages of second language acquisition. For instance, one
could expect English participants to incorrectly translate the Spanish
word \emph{botón} as \emph{bottom} instead of as its correct translation
\emph{button}, due to combined effect of the close phonological
similarity between \emph{botón} and \emph{bottom} along with the high
lexical frequency of \emph{bottom} relative to \emph{button}. To test
this prediction, we developed a lexical frequency-dependent measure of
cross-language phonological neighbourhood density, in which a neighbour
is counted only if it is higher frequency and is one phoneme apart from
the presented word. If the phonological neighbourhood density of the
target translation affects participants' performance negatively, this
would suggest that competitors in the native language affect recognition
of non-native words during recognition of foreign speech. We conducted a
series of three experiments to test these predictions.

@@

In each trial, participants listened to one audio file, wich contained a
single word. The audio files were the same ones used in child
experiments conducted in the Laboratori de Recerca en Infància of
Universitat Pompeu Fabra (Barcelona, Spain). These audio files were
recorded by a proficient Catalan-Spanish female bilingual from the
Metropolitan Area of Barcelona in a child-directed manner. Catalan and
Spanish words were recorded at 44,100 Hz in separate files in the same
session, and then de-noised using Audacity and normalised at peak
intensity using Praat (\citeproc{ref-broersma2021praat}{Broersma \&
Weenink, 2021}). The average duration of the Catalan audio files was
1.24 seconds (\emph{SD} = 0.19, Range = 0.8-1.58). The average duration
of the Spanish audio files was 1.16 seconds (\emph{SD} = 0.15, Range =
0.78-1.53).

For each word in the Catalan and Spanish lists pair, we defined three
predictors of interest: the lexical frequency of the correct English
translation (\emph{Frequency}), the phonological similarity between the
presented word (in Catalan or Spanish) and their correct English
translation (\emph{Similarity}), and the presented word's number of
cross-language phonological neighbours (\emph{CLPN}) in English.
\emph{Frequency} was included as a nuance predictor, under the
hypothesis that---keeping other predictors constant---participants would
translate higher-frequency words more accurately and faster than
lower-frequency words. Lexical frequencies of correct translations were
extracted from SUBTLEX-UK (\citeproc{ref-van2014subtlex}{Van Heuven et
al., 2014}), and transformed to Zipf scores. Words in the stimuli list
without a lexical frequency value were excluded from data analysis (2 in
the Catalan list, 5 in the Spanish list).

We calculated \emph{Similarity}, our main predictor of interest, by
computing the Levenshtein similarity between the X-SAMPA transcriptions
of each pair of translations using the \texttt{stringdist} R package
(\citeproc{ref-van2014stringdist}{van der Loo, 2014}). The Levenshtein
distance computes the edit distance between two character strings---in
this case, two phonological transcriptions---by counting the number of
additions, deletions, and substitutions necessary to make both strings
identical (\citeproc{ref-levenshtein1966binary}{Levenshtein, 1966}). We
divided this edit distance by the number of characters included in the
longest X-SAMPA transcription of the translation pair. This results in a
proportion score, in which values closer to zero correspond to lower
Levenshtein distances between phonological transcriptions (i.e., higher
similarity), and values closer to 1 correspond to higher Levenshtein
distances (i.e., lower similarity). This transformation accounts for the
fact that the Levenshtein distance tends to increase with the length of
the transcriptions. For interpretability, we subtracted this proportion
from one, so that values closer to one correspond to higher similarity
between phonological transcriptions, and values closer to zero
correspond to lower similarity between phonological transcriptions. For
example, the \emph{table} (\texttt{teIb@l})-\emph{mesa} (\texttt{mesa})
translation pair had a 17\% similarity, while the \emph{train}
(\texttt{trEIn})-\emph{tren} (\texttt{tREn}) translation pair had a 60\%
similarity. Table~\ref{tbl-stimuli} summarises the lexical frequency,
phonological neighbourhood density and phonological overlap of the words
included in the Catalan and the Spanish lists.

For each Catalan and Spanish word, we calculated the number of
\emph{CLPN} by counting the number of English words with same or higher
lexical frequency, and whose phonological transcription (in X-SAMPA
format) different in up to one phoneme from that of the presented
Catalan or Spanish word. Lexical frequencies and phonological
transcriptions were extracted from the multilingual database CLEARPOND
(\citeproc{ref-marian2012clearpond}{Marian et al., 2012})\footnote{{[}Phonological
  transcriptions in CLEARPOND were generated from eSPEAK,
  \url{http://espeak.sourceforge.net/}{]}}.

@@

The experiment was implemented online using Psychopy/Pavlovia
(\citeproc{ref-peirce2019psychopy2}{Peirce et al., 2019}). Participants
accessed the study from a link provided by Prolific or SONA and
completed the experiment from an internet browser (Chrome or Mozilla).
After giving their consent for participating, participants answered a
series of questions about their demographic status, their language
background, and the set up they were using for completing the study.
Then, participants completed the experimental task. Participants were
informed that they would listen to a series of pre-recorded words in
either Catalan or Spanish (English participants) or Catalan (Spanish
participants). They were instructed to listen to each word, guess its
meaning in English (English participants) or Spanish (Spanish
participants), and type their answer as soon as possible. English
participants were randomly assigned to the Catalan or Spanish lists.
Participants in the Catalan list were presented with 83 trials, and
participants in the Spanish list were presented with 99 trials.

Each trial started with a yellow fixation point presented during one
second on the centre of the screen over a black background. After one
second, the audio started playing while the dot remained being displayed
until the audio ended. Upon the offset of the fixation point and audio,
participants were prompted to write their answer by a
\texttt{\textgreater{}} symbol. Typed letters were displayed in the
screen in real time to provide visual feed-back to participants.
Participants were allowed to correct their answer. Then, participants
pressed the \texttt{RETURN/ENTER} key to confirm their answer and start
and new trial.

@@

\caption{Schematic representation of a trial in the experimental task.
The trial stars with the presentation of a fixation point in the centre
of the screen (yellow dot). After 1,000 ms, the auditory word was
presented while the fixation point remained on screen. After the offset
of the audio, the fixation point disappeared and a visual prompt
(\texttt{\textgreater{}}) was presented. Participants then wrote their
and clicked \texttt{RETURN}, making the end of the trial.}

@@

After data collection, participants' answers were manually coded into
the following categories: \emph{Correct}, \emph{Typo}, \emph{Wrong},
\emph{False friend}, \emph{Other}. A response was coded as
\emph{Correct} if the provided string of characters was identical to the
orthographic form of the correct translation. A response was coded as
\emph{Typo} if the participant provided a string of characters only one
edit distance (addition, deletion, or substitution) apart from the
orthographic form of the correct translation (e.g., ``pengiun'' instead
of ``penguin''), as long as the response did not correspond to a
distinct English word. A response was coded as \emph{False friend} if
the participant provided a phonologically similar incorrect translation.
Responses not meeting the criteria for previous categories were labelled
as \emph{Wrong} or \emph{Other} (see Data analysis section for more
details). Both \emph{Correct} and \emph{Typo} responses were considered
as correct, while \emph{Wrong} and \emph{False friend} responses were
considered as incorrect. \emph{Other} responses were excluded from data
analysis. Trials in which participants took longer than 10 seconds to
respond were also excluded. Participants contributed a total of 9,152
valid trials (5,694 in Catalan, 3,458 in Spanish). The task took
approximately 15 minutes to be completed.

@@

We modelled the probability of participants guessing the correct
translation of each presented word using a generalised multilevel
Bayesian regression model with a Bernoulli logit link distribution. We
included as fixed effects the intercept, the main effects of
\emph{Frequency}, \emph{Similarity}, and \emph{CLPN}, and the two-way
interaction between \emph{Similarity} and \emph{CLPN}. We also included
participant-level random intercepts and slopes for the the main effects
and the interaction. Eq. 1 shows a formal description of the model.

@@

We collected data for a total of 6,446 trials completed by 72 unique
participants. Of those trials, 2,988 were provided by 36 unique
participants who listened to Catalan words, and 3,458 trials were
provided by 36 unique participants who listened to Spanish words. We
excluded trials in which participants did not enter any text (\emph{n} =
72), in which a response in a language other than English was provided
(e.g., ``\texttt{agua}'', \emph{n} = 51), in which participants did not
provide a whole word (e.g., ``\texttt{f}'', \emph{n} = 5), and in which
participants added comments to the experimenter (e.g.,
``\texttt{unsure}'', \emph{n} = 13). In addition, we excluded data from
participants that self-rated their oral and/or written skills in Catalan
and Spanish, or any other second language as four or higher in a
five-point scale (\emph{n} = 2), were diagnosed with a language
(\emph{n} = 1), or did not contribute more than 80\% of valid trials
(\emph{n} = 9).

@@

\caption{\label{fig-epreds-2}Posterior model-predicted mean accuracy in
Experiment 1. Predictions were generated from 4,000 posterior samples,
extracted for different values of \emph{CLPN} (0, 2, 4, 8, 12) and
\emph{Similarity} (1-100). Predictions are plotted separately for
English participants translating Catalan words, and for English
participants translating Spanish words. Lines indicate mean predictions,
and intervals indicate 95\%, 89\%, 78\%, 67\%, and 50\% credible
intervals (CrI).}

>>>>

When some German speakers listened to the song \emph{The Power} by SNAP!
(\emph{Coyote Ugly}, 2000), many of them misheard the line ``I've got
the power'' as \emph{Agatha Bauer}. When Spanish speakers listen to the
line ``Circumvent your thick ego'' from the song \emph{Pictures} by
System of a Down (\emph{Steal this Album!}, 2002), they tend to hear
\emph{Sácame de aquí, King-Kong} {[}take me out of here, King-Kong{]}.
Outrageous as these examples might sound, the reader may know a few
cases in their own native language. This auditory illusion is common
across languages and cultures
(\citeproc{ref-dembeck2015oberflachenubersetzung}{Dembeck, 2015};
\citeproc{ref-efimova2018homophonic}{Efimova et al., 2018}) and can feel
quite real, and often inevitable. In Japanese, this phenomenon takes the
name \emph{Soramimi} (lit. ``empty ear''). Soramimi is a particular case
of homophonic translation in which words or phrases that appear in the
lyrics of a song in one language are translated into similar-sounding
words and phrases in a different language, without necessarily
preserving the meaning (\citeproc{ref-gasparov2006semen}{Gasparov,
2006}). Homophonic translation also occurs outside the context of music.
For instance, it is often intentionally employed by translators to
preserve the meter and ``sound structure'' (e.g., rhymes) of an original
text in the language of translation (e.g.,
\citeproc{ref-levick2016translating}{Levick, 2016};
\citeproc{ref-pilshchikov2016semiotics}{Pilshchikov, 2016})\footnote{The
  rationale behind this translation approach is to ``sacrifice{[}s{]}
  fidelity to the image rather than the melodiousness of verse''
  (\citeproc{ref-briusov1911ot}{Briusov, 1911}). For instance, Edith
  Grossman's translation of \emph{El Quijote}
  (\citeproc{ref-cervantes}{Cervantes \& Grossman, 2002}) present a
  similar case, in which formal aspects of the original text in Spanish
  are favoured to some degree over equivalence to English, even
  sometimes keeping Spanish words from the original text in the English
  translation (e.g., \emph{Señor}, \emph{ínsula}).}. Although homophonic
translation is mostly seen as a perceptual curiosity or as a literary
figure, little attention has been paid to homophonic translation as a
spontanous psycholinguistic event.

Otake (\citeproc{ref-otake2007interlingual}{2007}) analysed 194
instances of Soramimi, broadcasted between 1992 and 2007 by the TV show
\emph{Soramimi hour}, in which examples of Soramimi were presented the
the audience for comedy purposes. These instances consisted of
homophonic translations of English song lyrics to words and phrases in
Japanese. The vast majority of the reported instances were phrases
(96\%) and few were single words (4\%). There was relatively high
variability in the degree to which phonetic features of the presented
English lyrics were preserved, with some resulting Japanese words or
phrases sharing high overlap with the original English strings of
sounds, and some sharing very little overlap. An analysis of the few
instances of homophonic translations of single words revealed three main
phonological processes that could explain how Japanese listeners
reconstructed the English input to generate the Japanese words:
insertion (e.g., \emph{cry} /\textipa{kraI}/ to \emph{kurai},
/\textipa{kuRai}/ {[}dark{]}), deletion (e.g., \emph{go},
/\textipa{goU}/ to \emph{go}, /go/ {[}go{]}, and alternation (e.g.,
\emph{low}, /\textipa{loU}/ to \emph{rou}, /\textipa{Roo}/ {[}wax{]}).
This suggests that the reported homophonic translations can be
explained, to some extent, as the result of the Japanese listeners
accommodating the strings of English sounds to the Japanese phoneme
inventory and Japanese phonotactics
(\citeproc{ref-dupoux1999epenthetic}{Dupoux et al., 1999};
\citeproc{ref-peperkamp2008perceptual}{Peperkamp et al., 2008}).

In the present work, we investigated the interplay between phonological
similarity and phonological neighbourhood density to examine how
listeners may activate lexical representations in their native language
when listening to an unfamiliar language. Although homophonic
translations do not necessarily take place with the preservation of
meaning---which results in outrageous mistranslations like the
aforementioned ones---homophonic translations sometimes may lead to
correct translations. For example, imagine an English native listening
to Dutch for the first time. This listener may encounter the word
/\textipa{"pINgUIn}/ (\emph{pinguin}), which sounds similar to its
English translation /\textipa{"pENgwIn}/ (penguin). Given enough
phonological similarity between the non-native word and its translation
in the native language, a naïve listener may succeed at word
comprehension when listening to an unfamiliar language. Such
phonological similarity between translation equivalents---known as
\emph{cognateness}---is common across many languages, and is often due
to typological closeness and/or socio-historical events involving the
speakers of these languages (e.g., migration, social contact). For
example, Romance languages such as Spanish and Catalan share many
form-similar translation equivalents
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and
Catalan, respectively). Given no prior knowledge of either language, a
Spanish native speaker is likely to be much more successful at correctly
translating Catalan \emph{porta} than English \emph{door}, due to the
phonological and orthographic similarity of the former to the equivalent
in their native language.

When comparing the phonemic transcription of the French to that of its
Spanish translation /\textipa{"pweR.ta}/ (\emph{puerta}), one may be led
to infer that these two words do not have much overlap. But one thing to
note is that the voiced uvular fricative consonant /\textipa{K}/ and the
open-mid back rounded vowel /\textipa{O}/, which do not exist in Spanish
as phonemic categories, can be perceived as allophonic variations the
native (Spanish) phonemes /\textipa{R}/ and /o/, as described by
assimilation models of phoneme perception
(\citeproc{ref-best1988examination}{Best et al., 1988},
\citeproc{ref-best2001discrimination}{2001}). Although this phonetic
mismatch between the non-native word and its translation can have a
noticeable toll on comprehension Cutler et al.
(\citeproc{ref-cutler2004patterns}{2004}), the Spanish listener may
still be able to activate the phonological representation of
/\textipa{"pweR.ta}/ (\emph{puerta}) from hearing the word-form
/\textipa{"pOKt}/ (\emph{porte}), ultimately allowing them to access the
correct semantic representation. Native prosody also constrains the
occurrence of homophonic translations. For instance, Kentner
(\citeproc{ref-kentner2015rhythmic}{2015}) presented in the auditory
modality French-native and German-native participants with English songs
with lyrics. Homophonic translations followed language-specific
segmentation strategies. Most homophonic translations reported by
German-native participants resulted from inserting phrase and word
boundaries before stressed syllables (following the more frequent
trochaic pattern in German). On the contrary, most homophonic
translations reported by French-native participants resulted from
inserting phrase and word boundaries after stressed syllables, following
the more frequent iambic pattern in French.

To explore the lexical mechanisms triggered by unfamiliar speech
listening, we used a translation elicitation task in which participants
listened to words from an unfamiliar language, and then tried to guess
the words translations in their native language. We will henceforth
refer to the auditory-presented words heard by participants on each
trial as \emph{presented words}, and the correct translation for the
presented words as \emph{target translations}. By manipulating the
amount of phonological similarity between the presented words and their
target translations, we explored listeners' reliance on phonological
similarity to succeed in the task. Since participants were unfamiliar
with the presented language, they could only use phonological similarity
between the presented language and their native language to successfully
translate the words. We therefore predicted that participants'
performance should increase when the translation pairs are
phonologically more similar. We also predicted that there is a minimum
threshold of phonological similarity to be sufficient for translation.

We further investigated the effect of phonological competitors. Even if
the presented word and its target translation share high phonological
similarity, participants may still provide incorrect translations if the
presented word also shares high phonological similarity with other words
in the native language. Words with denser phonological neighbourhoods
are recognised more slowly and less accurately than words with sparser
phonological neighbourhoods (\citeproc{ref-dufour2003lexical}{Dufour \&
Peereman, 2003}; \citeproc{ref-goldinger1989priming}{Goldinger et al.,
1989}; \citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek,
1996}; \citeproc{ref-luce1990similarity}{Luce et al., 1990};
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). This is
especially true if such neighbours share higher phonological similarity
with the presented word, or are lexically more frequent than the target
word (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). Such
interference effect of phonological neighbourhood density has also been
reported across languages (\citeproc{ref-weber2004lexical}{Weber \&
Cutler, 2004}). We hypothesised that the size of the facilitatory role
of phonological similarity would be inversely proportional to the amount
of higher-frequency phonological neighbours of the presented word in the
target language. In the case of non-native speech recognition, the
presence of cross-linguistic pairs which are phonologically similar but
differ in meaning (e.g., false friends) may act as distractors during
lexical access, obstructing the selection of the appropriate target
translation in the listener's lexicon. For instance, Otwinowska \&
Szewczyk (\citeproc{ref-otwinowska2019more}{2019}) reported that false
friends were disadvantaged relative to non-cognates by Polish second
language learners, in contrast to cognates which were known better. It
is therefore important to investigate the joint effect of both cognates
and false friends when investigating the effect cross-linguistic
phonological similarity has on word recognition in a foreign language.
This could shed light on available strategies and challenges associated
with the early stages of second language acquisition. For instance, one
could expect English participants to incorrectly translate the Spanish
word \emph{botón} as \emph{bottom} instead of as its correct translation
\emph{button}, due to combined effect of the close phonological
similarity between \emph{botón} and \emph{bottom} along with the high
lexical frequency of \emph{bottom} relative to \emph{button}. To test
this prediction, we developed a lexical frequency-dependent measure of
cross-language phonological neighbourhood density, in which a neighbour
is counted only if it is higher frequency and is one phoneme apart from
the presented word. If the phonological neighbourhood density of the
target translation affects participants' performance negatively, this
would suggest that competitors in the native language affect recognition
of non-native words during recognition of foreign speech. We conducted a
series of three experiments to test these predictions.

@@

In each trial, participants listened to one audio file, which contained
a single word. The audio files were the same ones used in child
experiments conducted in the Laboratori de Recerca en Infància of
Universitat Pompeu Fabra (Barcelona, Spain). These audio files were
recorded by a proficient Catalan-Spanish female bilingual from the
Metropolitan Area of Barcelona in a child-directed manner. Catalan and
Spanish words were recorded at 44,100 Hz in separate files in the same
session, and then de-noised using Audacity and normalised at peak
intensity using Praat (\citeproc{ref-broersma2021praat}{Broersma \&
Weenink, 2021}). The average duration of the Catalan audio files was
1.24 seconds (\emph{SD} = 0.19, Range = 0.8-1.58). The average duration
of the Spanish audio files was 1.16 seconds (\emph{SD} = 0.15, Range =
0.78-1.53).

For each word in the Catalan and Spanish lists pair, we defined three
predictors of interest: the lexical frequency of the correct English
translation (\emph{Frequency}), the phonological similarity between the
presented word and their correct English translation
(\emph{Similarity}), and the presented word's number of cross-language
phonological neighbours (\emph{CLPN}) in English. \emph{Frequency} was
included as a nuance predictor, under the hypothesis that---keeping
other predictors constant---participants would provide correct responses
more often when the correct translations are high frequency words in
their native language. Lexical frequencies of correct translations were
extracted from SUBTLEX-UK (\citeproc{ref-van2014subtlex}{Van Heuven et
al., 2014}), and transformed to Zipf scores. Translations without a
lexical frequency value were excluded from data analysis (2 in the
Catalan list, 5 in the Spanish list).

We calculated \emph{Similarity}, our main predictor of interest, by
computing the Levenshtein similarity between the X-SAMPA transcriptions
of each pair of translations using the \texttt{stringdist} R package
(\citeproc{ref-van2014stringdist}{van der Loo, 2014}). The Levenshtein
distance computes the edit distance between two character strings---in
this case, two phonological transcriptions---by counting the number of
additions, deletions, and substitutions necessary to make both strings
identical (\citeproc{ref-levenshtein1966binary}{Levenshtein, 1966}). We
divided this edit distance by the number of characters in the longest
X-SAMPA transcription of the translation pair. This transformation
accounts for the fact that Levenshtein distance tends to increase with
the length of the transcriptions. For interpretability, we subtracted
this proportion from one, so that values closer to one correspond to
higher similarity between phonological transcriptions, and values closer
to zero correspond to lower similarity between phonological
transcriptions. For example, the \emph{table}
(\texttt{teIb@l})-\emph{mesa} (\texttt{mesa}) translation pair had 17\%
similarity, while the \emph{train} (\texttt{trEIn})-\emph{tren}
(\texttt{tREn}) translation pair had 60\% similarity.

For each Catalan and Spanish word, we calculated the number of
\emph{CLPN} by counting the number of English words with same or higher
lexical frequency, and whose phonological transcription (in X-SAMPA
format) differed in up to one phoneme from that of the presented Catalan
or Spanish word. Lexical frequencies and phonological transcriptions
were extracted from the multilingual database CLEARPOND
(\citeproc{ref-marian2012clearpond}{Marian et al., 2012})\footnote{{[}Phonological
  transcriptions in CLEARPOND were generated from eSPEAK,
  \url{http://espeak.sourceforge.net/}{]}}. Table~\ref{tbl-stimuli}
summarises the lexical frequency, phonological neighbourhood density and
phonological overlap of the words included in the Catalan and the
Spanish lists.

@@

The experiment was implemented online using Psychopy/Pavlovia
(\citeproc{ref-peirce2019psychopy2}{Peirce et al., 2019}). Participants
accessed the study from a link provided by Prolific or SONA and
completed the experiment from an internet browser (Chrome or Mozilla).
After giving their consent for participating, participants answered a
series of questions about their demographic status, their language
background, and the set up they were using for completing the study.
Then, participants completed the experimental task. Participants were
informed that they would listen to a series of pre-recorded words in
either Catalan or Spanish. They were instructed to listen to each word,
guess its meaning in English, and type their answer as soon as possible.
Participants were randomly assigned to the Catalan or Spanish lists.
Participants in the Catalan list were presented with 83 trials, and
participants in the Spanish list were presented with 99 trials.

Each trial started with a yellow fixation point presented during one
second on the centre of the screen over a black background. After one
second, the audio started playing while the dot remained being displayed
until the audio ended. Upon the offset of the fixation point and audio,
participants were prompted to write their answer by a
\texttt{\textgreater{}} symbol. Typed letters were displayed on the
screen in real time to provide visual feed-back to participants.
Participants were allowed to correct their answer. Then, participants
pressed the \texttt{RETURN/ENTER} key to confirm their answer and start
a new trial.

@@

\caption{Schematic representation of a trial in the experimental task.
The trial stars with the presentation of a fixation point in the centre
of the screen (yellow dot). After 1,000 ms, the auditory word was
presented while the fixation point remained on screen. After the offset
of the audio, the fixation point disappeared and a visual prompt
(\texttt{\textgreater{}}) was presented. Participants then wrote their
answer and clicked \texttt{RETURN}, making the end of the trial.}

@@

After data collection, participants' answers were manually coded into
the following categories: \emph{Correct}, \emph{Typo}, \emph{Wrong},
\emph{False friend}, \emph{Other}. A response was coded as
\emph{Correct} if the provided string of characters was identical to the
orthographic form of the correct translation. A response was coded as
\emph{Typo} if the participant provided a string of characters that was
only one edit distance (addition, deletion, or substitution) apart from
the orthographic form of the correct translation (e.g., ``pengiun''
instead of ``penguin''), as long as the response did not correspond to a
distinct English word. A response was coded as \emph{False friend} if
the participant's response was incorrect, but phonologically similar to
the presented word. Responses not meeting the criteria for previous
categories were labelled as \emph{Wrong} or \emph{Other} (see Data
analysis section for more details). Both \emph{Correct} and \emph{Typo}
responses were considered as correct, while \emph{Wrong} and \emph{False
friend} responses were considered as incorrect. \emph{Other} responses
were excluded from data analysis. Trials in which participants took
longer than 10 seconds to respond were also excluded. Participants
contributed a total of 5,206 valid trials (2,604 in Catalan, 2,602 in
Spanish). The task took approximately 15 minutes to complete.

@@

We modelled the probability of participants guessing the correct
translation of each presented word using a generalised multilevel
Bayesian regression model with a Bernoulli logit link distribution. We
included as fixed effects the intercept, the main effects of
\emph{Frequency}, \emph{Similarity}, and \emph{CLPN}, and the two-way
interaction between \emph{Similarity} and \emph{CLPN}. We also included
participant-level random intercepts and slopes for the main effects and
the interaction. Eq. 1 shows a formal description of the model.

@@

We collected data for a total of 6,446 trials completed by 72 unique
participants. We excluded trials in which participants did not enter any
text (\emph{n} = 72), in which a response in a language other than
English was provided (e.g., ``\texttt{agua}'', \emph{n} = 51), in which
participants did not provide a whole word (e.g., ``\texttt{f}'',
\emph{n} = 5), and in which participants added comments to the
experimenter (e.g., ``\texttt{unsure}'', \emph{n} = 13). In addition, we
excluded data from participants that self-rated their oral and/or
written skills in Catalan and Spanish, or any other second language as
four or higher in a five-point scale (\emph{n} = 2), were diagnosed with
a language (\emph{n} = 1), or did not contribute more than 80\% of valid
trials (\emph{n} = 9).

@@

\caption{\label{fig-epreds-2}Posterior model-predicted mean accuracy in
Experiment 1. Predictions were generated from 4,000 posterior samples,
extracted for different values of \emph{CLPN} (0, 2, 4, 8, 12) and
\emph{Similarity} (1-100). Lines indicate mean predictions, and
intervals indicate 95\%, 89\%, 78\%, 67\%, and 50\% credible intervals
(CrI).}
